PBJ: A Gnutella Inspired File Sharing System

### 2012-01-05 paper/paper.tex
updated testing section

Peer to peer (P2P) file sharing applications are some of the most well known and ubiquitous distributed systems today. These applications allow users of their system to share and download files between each other (usually) free of charge.  In the last fifteen years, file sharing has gone from virtually nonexistent to one of the largest sources of Internet traffic ~\cite{study}.  P2P system structure can vary wildly from highly centralized, in which a single server handles all requests, to highly decentralized, where there are no central elements and the user client programs work together to find results. Beyond basic topology, there are a host of design trade offs.
We were also interested in creating a system that would not be considered liable if illegal content were shared on it. This has been a problem for file sharing applications in the past, particularly Napster. PBJ is a distributed file system that has been designed to be stable, reliable, and able to adapt to changing needs of its users. Due to these design goals, we decided to loosely base the design of PBJ on gnutella, a highly distributed file sharing application. While our overall design is gnutella based, we looked at a range of famous file sharing systems, and several of them influenced our decision making during the design process.

\section{Related Work}
\subsection{Napster}
A defining characteristic of a P2P file sharing application is the kind of network topology it employs.  Napster, one of the first file sharing networks, was a highly centralized system.  All  online Napster users connected to a central server.  Each user would have a folder designated for sharing audio files (only mp3s were transferred on Napster).  This central server would handle a user request by searching for related files in the share folders of other users.  The central server would then report a list of relevant files and their locations that the user could choose to download.  Any files the requesting user chose to download would directly connect with the associated machine for actual file transfer. Because a given node has only to contact the central server, lookup in Napster can be done in constant time. ~\cite{superpeer}

@@

 \begin{figure}[tbh]
\centering
\includegraphics[width=0.5\textwidth]{362prez}
\caption{PBJ Network of Ultra Peers}
\end{figure}
The basic algorithm for connecting ultra nodes in the network builds an outward spiral of ultra peers.  While this alone would be wildly inefficient in terms of searching and stability, new ultra nodes also connect to ultra nodes deeper within the spiral.  This allows search requests to quickly move between otherwise distant ultra nodes. More specifically, the first ultra node created will be given an ID of 0, the second an ID of 1, etc.  When an ultra node is created with ID Y, it will attempt to connect to all nodes $X, 0 < X < Y$, where $\log_2 (|Y - X|) = Z , Z$ being an even integer.  So for example, node 16 would connect with node 15 $(16 - 15 = 1 = 2^0)$, node 12 $(16 - 12 = 4 = 2^2)$, and node 0 $(16 - 0 = 16 = 2^4)$.

A clear benefit of this structure is that the PBJ network has a high level of redundancy in the connections.  This means that the nodes are connected in such a way that it is easy for search requests to reach most, if not all, of the nodes on the network in a few hops.  Currently, the process of actually adding ultra peers to the network and telling them which other ultra peers to connect to is handled through the gateway. The main job of the gateway is to make decisions about how to add a new user node to the network. The gateway keeps track of the status of all the ultrapeers currently in the system.  When a new node tries to join the system, the gateway makes several decisions based on the current status of the PBJ network.  First, the gateway decides if the new node will become an ultra node.  If so, it will give the new ultra node a list of other ultra nodes to connect to.  If the new node is not an ultra node, the gateway will provide information on the appropriate ultra peer sub network to join. These decisions by the gateway essentially dictate how the network is built. We devised two separate algorithms for building the network; the one we actually implemented is described here and the other is detailed in the section “Future Work”.  As implemented, the first node to enter the network becomes ultra node 0.  If, for example, there are three nodes per ultra node, the next three nodes to connect would become node 0’s sub network.  The fifth node to join the network would become ultra node 1 and would connect to ultra node 0. In this scheme of building the network, each ultra node’s sub network is filled before the next ultra node is created.  This algorithm ensures that the graph has the smallest number of ultra nodes possible in a system and all but potentially the last ultra node have filled sub networks.

@@

\section{Results}
All of the PBJ network communication is facilitated through the use of the Flask python module. Flask ~\cite{flask} is a web microframework that is designed to simplify the development of Python web applications. HTTP was chosen as a protocol over TCP due to its reliability, simplicity, and flexibility. Flask trivializes the implementation of custom HTTP responses. Every node, as well as the gateway, is running an instance of a Flask server. The server is configured to run on port 5000 as a default, but this could be easily changed.
Flask was not available as an installed Python module on the lab machines that we were planning on using for our testing, and we did not have administrative privileges to install it, so we needed a way in which we could install and use Flask on every lab machine without modifying any system files. The solution was to use virtualenv ~\cite{virtual}, which is a tool used to create virtual Python environments. A virtual Python environment can be created in a users home folder on a system using virtualenv, and then that user has permission to modify that environment however they wish. In this way, we were able to deploy a Python environment with the Flask module and its dependencies installed on every lab machine.

A concern of ours when laying out the specification of how nodes would communicate was the case in which a node would have to communicate across a firewall. Firewalls add a great deal of complexity when designing a distributed system that is intended to be accessed by a large number of computers under diverse network constraints. Our ultimate decision was that since our project is academic in nature and limited in resources, we were going to ignore the cases in which communication over port 5000 is unavailable to a node.

\subsection{Test Plan}
In order to claim that we have designed a stable and scalable system with confidence, we needed to design a solid testing plan. In the design phase of our project, we thoroughly discussed the use of Condor, a HTC application that can deploy processes to a large number of computers, to test our network stability and search algorithms for large numbers of nodes. Unfortunately, the computers on Clemson’s Condor network are equipped with Windows Firewall, which blocks the access of ports that have not been explicitly granted permission. As previously discussed, we will be ignoring the cases in which our nodes are behind firewalls, so this test strategy was inadequate. Instead, we devised a test strategy in which we wrote a script to connect every available lab machine we had to our network.

The end result was a test environment in which over 70 search-able nodes with unique files were connected to our network. Although this is a considerable amount of nodes, we found that it was not sufficient to fully test the scaling and performance of our search algorithm. We used this large network of connected nodes primarily to test the functionality and reliability of our system.
The main goals of testing the gateway server was twofold. First, correct network topology was to be maintained when new nodes were added.  This could easily be checked by feeding the gateway a large number of pseudo nodes to verify that the intended topology could be maintained for any size network. Furthermore, an applet was developed to display an image representing the network graph, making it easier to see node to node connections. The second goal of gateway testing was to verify that any dropped node notifications were being handled correctly, which involved updating the network structure for addition of future nodes. This is further discussed later when testing node functions.

Testing node functionality proved to be much more in depth. For this, we developed a series of repeatable test cases that the network should be able to handle. These test cases included: initiating a search as an ultra node, initiating a search as a node, searching from one end of the network to the other, searching from the middle of the network to either end, and downloading a search result. We also monitored the gateway to ensure that the network structure was maintained and rebuilt after all node disconnection scenarios: dropping a node from the beginning, middle, and end of the network, as well as dropping an ultra node from the beginning, middle, and end of the network.
We repeated these test cases after every change we made to ensure that PBJ was stable and working properly.
\section{Analysis}

@@

Our results show that there is a high correlation between network structure and the number of hops required to reach all the nodes on the system.  Looking at the average number of hops to reach every ultra peer in a network, several trends become apparent.  First, every time a ultra peer with an even power of two connects to ultra peer 0, (ultra peers 4, 16, 64, and 256) the average number of hops to traverse the network decreases by approximately one hop.  Once this drop has occurred, as new ultra peers are added to the network, the average number of hops grows at a roughly linear rate to about one more hop on average than the previous peak.  Notice that the slope between peaks is decreasing with an increasing number of nodes.  This implies that as PBJ gets more and more ultra peers, the addition of ultra peers will have less of an impact on average number of hops to find a file.

>>>>

Peer to peer (P2P) file sharing applications are some of the most well known and ubiquitous distributed systems today. These applications allow users of their system to share and download files between each other (usually) free of charge.  In the last fifteen years, file sharing has gone from virtually nonexistent to one of the largest sources of Internet traffic ~\cite{study}.  P2P system structure can vary wildly from highly centralized, in which a single server handles all requests, to highly decentralized, where there are no central elements and the user client programs work together to find results. Beyond basic topology, there are a host of design trade offs.
We were interested in creating a system that would not be considered liable if illegal content were shared on it. This has been a problem for file sharing applications in the past, particularly Napster. PBJ is a distributed file system that has been designed to be stable, reliable, and able to adapt to changing needs of its users. Due to these design goals, we decided to loosely base the design of PBJ on gnutella, a highly distributed file sharing application. While our overall design is gnutella based, we looked at a range of famous file sharing systems, and several of them influenced our decision making during the design process.

\section{Related Work}
\subsection{Napster}
A defining characteristic of a P2P file sharing application is the kind of network topology it employs.  Napster, one of the first file sharing networks, was a highly centralized system.  All Napster users connected to a central server.  Each user would have a folder designated for sharing audio files (only mp3s were transferred on Napster).  This central server would handle a user request by searching for related files in the share folders of other users.  The central server would then report a list of relevant files and their locations that the user could choose to download.  Any files the requesting user chose to download would directly connect with the associated machine for actual file transfer. Because a given node has only to contact the central server, lookup in Napster can be done in constant time. ~\cite{superpeer}

@@

 \begin{figure}[tbh]
\centering
\includegraphics[width=0.5\textwidth]{362prez}
\caption{PBJ Network of Ultra Peers}
\end{figure}
The basic algorithm for connecting ultra nodes in the network can be visualized as an outward spiral of ultra peers.  While this alone would be wildly inefficient in terms of searching and stability, new ultra nodes also connect to ultra nodes deeper within the spiral.  This allows search requests to quickly move between otherwise distant ultra nodes. More specifically, the first ultra node created will be given an ID of 0, the second an ID of 1, etc.  When an ultra node is created with ID Y, it will attempt to connect to all nodes $X, 0 < X < Y$, where $\log_2 (|Y - X|) = Z , Z$ being an even integer.  So for example, node 16 would connect with node 15 $(16 - 15 = 1 = 2^0)$, node 12 $(16 - 12 = 4 = 2^2)$, and node 0 $(16 - 0 = 16 = 2^4)$.

A clear benefit of this structure is that the PBJ network has a high level of redundancy in the connections.  This means that the nodes are connected in such a way that it is easy for search requests to reach most, if not all, of the nodes on the network in a few hops.  Currently, the process of actually adding ultra peers to the network and telling them which other ultra peers to connect to is handled through the gateway. The main job of the gateway is to make decisions about how to add a new user node to the network. The gateway keeps track of the status of all the ultrapeers currently in the system.  When a new node tries to join the system, the gateway makes several decisions based on the current status of the PBJ network.  First, the gateway decides if the new node will become an ultra node.  If so, it will give the new ultra node a list of other ultra nodes to connect to.  If the new node is not an ultra node, the gateway will provide information on the appropriate ultra node sub network to join. These decisions by the gateway essentially dictate how the network is built. We devised two separate algorithms for building the network; the one we actually implemented is described here and the other is detailed in the section “Future Work”.  As implemented, the first node to enter the network becomes ultra node 0.  If, for example, there are three nodes per ultra node, the next three nodes to connect would become node 0’s sub network.  The fifth node to join the network would become ultra node 1 and would connect to ultra node 0. In this scheme of building the network, each ultra node’s sub network is filled before the next ultra node is created.  This algorithm ensures that the graph has the smallest number of ultra nodes possible in a system and all but potentially the last ultra node have filled sub networks.

@@

\section{Results}
All of the PBJ network communication is facilitated through the use of the Flask python module. Flask ~\cite{flask} is a web microframework that is designed to simplify the development of Python web applications. HTTP was chosen as a protocol over TCP due to its reliability, simplicity, and flexibility. Flask trivializes the implementation of custom HTTP responses. Every node, as well as the gateway, is running an instance of a Flask server. The instance of the Flask server is told to use on a random port that is not currently in use.
Flask was not available as an installed Python module on the machines that we were planning on using for our testing, and we did not have administrative privileges to install it, so we needed a way in which we could install and use Flask on every lab machine without modifying any system files. The solution was to use virtualenv ~\cite{virtual}, which is a tool used to create virtual Python environments. A virtual Python environment can be created in a users home folder on a system using virtualenv, and then that user has permission to modify that environment however they wish. In this way, we were able to deploy a Python environment with the Flask module and its dependencies installed on every lab machine.

A concern of ours when laying out the specification of how nodes would communicate was the case in which a node would have to communicate across a firewall. Firewalls add a great deal of complexity when designing a distributed system that is intended to be accessed by a large number of computers under diverse network constraints. Our ultimate decision was that since our project is academic in nature and limited in resources, we were going to ignore the cases in which communication over the allocated port is unavailable to a node.

\subsection{Test Plan}
In order to claim that we have designed a stable and scalable system with confidence, we needed to design a solid testing plan. In the design phase of our project, we thoroughly discussed the use of Condor, a HTC application that can deploy processes to a large number of computers, to test our network stability and search algorithms for large numbers of nodes. Unfortunately, the computers on Clemson’s Condor network are equipped with Windows Firewall, which blocks the access of ports that have not been explicitly granted permission. As previously discussed, we will be ignoring the cases in which our nodes are behind firewalls, so this test strategy was inadequate. Instead, we requested and were granted permission to perform our testing on Clemson's Palmetto cluster. Our account on the cluser utilized a shared network drive between nodes, so we had to write a script that would create a unique share file in a unique share directory. Each node was to have a single shared file for consistency. The end result was a test environment in which 300 search-able nodes with unique files were connected to our network.

We decided to focus our testing on the search performance and scalability of our network structure. To do this, we eliminated the sub networks so that every new node was designated as an ultra node. We also added additional functionality to the nodes to allow the gateway to tell a node to initiate a search. We designated ultra node 0 to be the searching node for our testing. The gateway was modified so that it initiated a search from node 0 upon the addition of a new node to the network. Node 0 then was to log the time the search took for each result, as well as how many hops the request made until each result was found. Nodes were added to the network at intervals of one second, which gave us 300 test results that describe the scalability of our system.

%can we remove this paragraph? its pretty much fluff/no longer relevant
%but i would feel bad deleting a whole paragraph without consulting
The main goals of testing the gateway server was twofold. First, correct network topology was to be maintained when new nodes were added.  This could easily be checked by feeding the gateway a large number of pseudo nodes to verify that the intended topology could be maintained for any size network. Furthermore, an applet was developed to display an image representing the network graph, making it easier to see node to node connections. The second goal of gateway testing was to verify that any dropped node notifications were being handled correctly, which involved updating the network structure for addition of future nodes. This is further discussed later when testing node functions.

%our we going to talk about our other "testing" or just the palmetto stuff? if the latter, del this too
Testing node functionality proved to be much more in depth. For this, we developed a series of repeatable test cases that the network should be able to handle. These test cases included: initiating a search as an ultra node, initiating a search as a node, searching from one end of the network to the other, searching from the middle of the network to either end, and downloading a search result. We also monitored the gateway to ensure that the network structure was maintained and rebuilt after all node disconnection scenarios: dropping a node from the beginning, middle, and end of the network, as well as dropping an ultra node from the beginning, middle, and end of the network.
We repeated these test cases after every change we made to ensure that PBJ was stable and working properly.
\section{Analysis}

@@

Our results show that there is a high correlation between network structure and the number of hops required to reach all the nodes on the system.  Looking at the average number of hops to reach every ultra peer in a network, several trends become apparent.  First, every time an ultra peer with an even power of two connects to ultra peer 0, (ultra peers 4, 16, 64, and 256) the average number of hops to traverse the network decreases by approximately one hop.  Once this drop has occurred, as new ultra peers are added to the network, the average number of hops grows at a roughly linear rate to about one more hop on average than the previous peak.  Notice that the slope between peaks is decreasing with an increasing number of nodes.  This implies that as PBJ gets more and more ultra peers, the addition of ultra peers will have less of an impact on average number of hops to find a file.
